An Analysis of Wind Energy Potential in the North Sea

### 2024-01-29 paper/paper.tex
updated the paper tex

\begin{abstract}
%\textcolor{magenta}{
%Put your abstract here. Abstracts typically start with a sentence motivating why the subject is interesting. Then mention the data, methodology or methods you are working with, and describe results.}
In the face of climate change, it is widely agreed that the energy production has to rely on more sustainable and renewable forms of harnessing energy. Offshore wind turbine parks play a crucial role in increasing the share of green energy. This paper explores probabilistic methods of assessing the wind energy potential and potential trends of data collected on Helgoland by considering wind speeds as a Weibull distributed random variable. Further, for forecasting, the monthly expected wind power density is extrapolated using a Gaussian process regression model.
\end{abstract}

@@

The wind speed data used in this paper was collected by \href{https://www.dwd.de}{DWD} on \href{https://www.openstreetmap.org/?mlat=54.1750&mlon=7.8920#map=15/54.1750/7.8920}{Helgoland}, spanning the years from 1996 to 2022. The wind speed is recorded at a temporal resolution of ten-minutes\footnote{Averaged over the last ten-minute interval.}, with the anemometer being $4.38 \mathrm{m}$ above ground. Averaged over the whole measurement period, the data has a yearly mean completeness of about $91\%$. Notably, the year 2019 is missing $88\%$ of its measurements. We consider this, and a noticeable drop in the yearly wind speed mean and variance of the following years, as evidence that the measurement process was changed, and therefore discard them. Notably, the years 1996 to 1998 have a completeness of about $3\%, 85\%$ and $84\%$, respectively. For preventing a distorting effect on the Weibull distribution estimations, we also did not consider these years for our analysis, thereby improving the yearly mean completeness to $99\%$. For handling outliers, we dropped measurements exhibiting a z-score $> 8$.

@@

\subsection{Fitting of the Weibull Distribution and Trend Analysis}\label{sec:weibull}
Based on the observed wind data, we model the wind speed probability distribution both on a monthly and yearly basis.
Whilst there are several other models, like the three-parameter Weibull distribution or the normal distribution for fitting wind speed frequencies, the most prominent one is the two-parameter Weibull distribution
\cite{review, statanalysis, mohammadi2016assessing}.
For the scale parameter $\lambda >0$ and shape parameter $\beta >0$, the probability density function (PDF) and cumulative density function (CDF) are given by
\begin{align}
    \label{eq:pdf}
    p(v) &=
    \begin{cases}
        \frac{\beta}{\lambda} \left( \frac{v}{\beta}\right)^{\beta-1} 
        \exp{ \left( -\left(\frac{v}{\lambda}\right)^\beta \right)} & \text{  for } v>0 \\
        0 & \text{ else}
    \end{cases} \\
    \label{eq:cdf}
    P(v)&=
    \begin{cases}
        1- \exp{\left(-\left(\frac{v}{\lambda}\right)^\beta \right)} & \text{  for } v>0 \\
        0 & \text{ else}
    \end{cases}.
\end{align}
Via the gamma function, $\Gamma(z)= \int_{0}^{\infty}t^{z-1} \mathrm{e}^{-t}dt$, 
$k$-th raw moment and variance of a random variable $V$ with Weibull distributions can be expressed as follows: 
\begin{align}  
    \mathrm{E}[V^k] &= \lambda^k \Gamma \left( 1 + \frac{k}{\beta} \right) \label{eq:n-raw-moment} \\
    \mathrm{Var}[V] &= \lambda^2 \left[  \Gamma \left( 1 + \frac{2}{\beta}\right) \label{eq:var}
    - \Gamma^2 \left( 1 + \frac{1}{\beta}\right)\right].
\end{align}

@@

\subsection{Gaussian Process Regression}\label{sec:gp}
For forecasting, we employ a Gaussian process regression (GPR) model to perform a time-series regression on the monthly estimated power density and extrapolate future values. Thereby, we assume that the wind speed data $Y \in \mathbb{R}^N$ is from a latent function $f: \mathbb{R}^N \rightarrow \mathbb{R}^N$ with some added measurement noise: $y = f(x) + \epsilon$. The GPR framework consists of a Gaussian process as prior, multivariate Gaussian likelihood and Gaussian process posterior conditioned on the data 
\cite{rasmussen-williams-gp}:
\begin{align}
    \label{eq:gp_prior}
    p(f) &= \mathcal{GP}(f; 0, k), \\
    \label{eq:gp_likelihood}
    p(y \mid f) &= \mathcal{N}(Y; f, \sigma^2 I), \\
    \label{eq:gp_posterior}
    p(f \mid Y) &= \mathcal{GP}(f; m', k') \propto p(Y \mid f) \cdot p(f).
\end{align}
The resulting posterior (and prior) distribution collapses into a multivariate normal distribution when evaluate at some new locations $X' \in \mathbb{R}^M$.

@@

The least-squares linear regression of the yearly parameters can be seen in Figure \ref{fig:lin_reg}.
It shows a decrease of the parameter $\lambda$ of 
$-0.059\frac{m}{s} \frac{1}{\mathrm{yr}}$ with an RMSE of $0.42\frac{m}{s}$. 
For $\beta$, the slope is $-0.002 \frac{1}{\mathrm{yr}}$, with an RMSE of $0.07  \frac{m}{s}$.
At the $5\%$ level of the two-sided permutation test, the null hypothesis that there is no linear trend in 
the development of the parameter $\lambda$ can be rejected (p-value$=0.002$), so the observation is statistically significant. 
For the parameter $\beta$, however, the p-value is $0.371$, so we cannot reject the null hypothesis.
Assuming the trend observed in $\lambda$ and the empiric mean $\bar \beta=2.28$ for $\beta$, we derive the following trends in other quantities:
As the mean  and standard deviation are proportional to $ \lambda$, we can infer a linear trend of $-0.052 \frac{m}{s}  \frac{1}{\mathrm{yr}}$
and  $-0.024 \frac{m}{s} \frac{1}{\mathrm{yr}}$, respectively. 
%Further, the probabilities of wind speed below $ 3 \frac{m}{s}$ and above $22 \frac{m}{s}$ are especially relevant (Section \ref{sec:power-density}):
%Over the entire $22$ year period, the estimated probabilities change (in a non-linear fashion) from $6.46 \%$ to $8.15 \%$ and from $0.17 \%$ to $0.03 \%$, respectively. 
Further, under these assumptions, the mean power density 
of the mean wind speed decreases (non-linearly) from 1999 to 2018  by $30\%$.

@@

%For assessing the viability of wind turbine (parks), one might also be interested in the weekly or daily values  and, even more importantly, in a forecast. Hence, for interpolating the daily expected power density, we employ a Gaussian Process Regression model to perform a time-series regression. In addition to interpolating, we also use the same model for extrapolation and thereby making predictions about the future daily expected power density.
The first step for conducting a GP regression is to decide on a kernel $k$ for the prior \eqref{eq:gp_prior}. By analysing the plotted raw monthly expected power density, we identify a yearly periodic trend and a slight global downward trend. The former can be captured by the Exponential-Sine-Squared kernel (ES) \cite{MacKay1998IntroductionTG}. Supported by the empirical data, we choose a periodicity of $365 \, \text{days}$ and an amplitude of $450 \, \frac{\textrm{W}}{\text{m}^2}$. For the length scale, we settle for a rather small value ($< 2 \, $ days) in order to capture local patterns in the periodicity more accurately. For allowing the amplitude to slowly decay over long periods of time, we multiply the ES kernel by a very smooth (length scale $> 50 \, \text{years}$) Squared-Exponential (SE) kernel. For reproducing the global downwards trend observed in section \ref{sec:trend}, we also choose a very smooth Squared-Exponential kernel with an output scale of $100 \, \frac{\textrm{W}}{\text{m}^2}$. For accounting for the monthly midterm trend, we add a Rational-Quadratic kernel \cite{rasmussen-williams-gp} with a scale of $10 \, \frac{\textrm{W}}{\text{m}^2}$ and a length scale of one month. Finally, the additive kernel, consisting of a White-Noise (WN) and a Matern kernel (MA) \cite{abramowitz1968handbook}, is responsible for explaining measurement noise in the data and thus also prevents overfitting. Thereby, the kernel has the form
$$k = \underbrace{\text{ES} \cdot \text{SE}}_{\text{yearly trend}} + \underbrace{\text{RQ}}_\text{monthly trend} + \underbrace{\text{ES}}_\text{longterm trend} + \underbrace{\text{WN} + \text{MA}}_{\text{noise}}.$$
By drawing samples\footnote{with the mean of the data added.} $f_i \sim \mathcal{GP}(f; 0, k)$ from the prior distribution \eqref{eq:gp_prior}, we gauge how accurately our kernel choice fits to the given data. The results are shown in Figure \ref{fig:gp-samples}. We can observe that the amplitude and periodicity of some of the samples mirrors the data's in an accurate manner, supporting our choice of hyperparameters. After conditioning the Gaussian Process prior on the normalised data, we obtain the posterior distribution $\mathcal{GP}(f; m', k')$ \eqref{eq:gp_posterior}. The evaluated posterior mean and standard deviation can be found in Figure \ref{fig:gp-pred}. We can observe that the mean function is following the periodicity of the ground truth without overfitting extreme power densities. Furthermore, the uncertainty of the model (standard deviation) increases gradually the further the prediction is from the data. Notably, the model also captures the global downward trend. When comparing the prediction to the test data in the years 2017 and 2018, we can observe that the prediction fits the ground truth accurately and all data points are well within one standard deviation of the mean.

@@

However, there are some shortcomings. One of the inherent limitations of the GPR is that it yields functions mapping onto $\R$, while the power density is only limited to $\R_+$. 
%Further, we conducted an interpolation of daily values based on monthly data, which does not take the inter-day variability into account. For the purpose of extrapolating daily power densities, 
%it might be advisable in further work to rely on daily data to begin with, complicating the choice of the kernel.  
Another important limitation is that our data was recorded at a height of $4.38 \mathrm{m}$, whereas the real height of wind turbines can exceed $100 \mathrm{m}$. One commonly applied rule of thumb to mitigate this issue
is multiplying the wind speeds by the factor 
$\left(\frac{h}{4.38 \mathrm{m}} \right)^{0.14}$ or the power density by its cube, 
where $h$ is the  proposed hub height of a turbine \citep{statanalysis}. However, this is far from exact, as the winds at different heights could be subject to different circulations. 
Furthermore, we did not take the direction of the wind into account and made the simplification that wind turbines are always perfectly aligned. 
Also, for the \eqref{eq:power}, we assumed a constant air pressure rather than considering the monthly changing pressure.
Lastly, with the use of the maximum-likelihood method, we assumed the independence of measurements. While this is generally not given, we were still able to produce adequate estimations.

\section*{Contribution Statement}
David Voigt devised the wind power density and conducted the Gaussian process regression. 
Gwendolyn Neitzel was mainly concerned with the Weibull model fitting and the trend analysis.
Mohammad Fadel Berakdar focused on the data preprocessing. Alireza Yahyanejad worked on data loading.
All authors were involved in data exploration and jointly wrote the report. 
%\textcolor{magenta}{Explain here, in one sentence per person, what each group member contributed. For example, you could write: Max Mustermann collected and prepared data. Gabi Musterfrau and John Doe performed the data analysis. Jane Doe produced visualizations. All authors will jointly write the text of the report. Note that you, as a group, are collectively responsible for the report. Your contributions should be roughly equal in amount and difficulty.}

>>>>

\begin{abstract}
%\textcolor{magenta}{
%Put your abstract here. Abstracts typically start with a sentence motivating why the subject is interesting. Then mention the data, methodology or methods you are working with, and describe results.}
In the face of climate change, it is widely agreed that energy production has to rely on more sustainable and renewable forms of harnessing energy. Offshore wind turbine parks play a crucial role in increasing the share of green energy. This paper explores probabilistic methods of assessing the wind energy potential and potential trends of data collected on Helgoland by considering wind speeds as a Weibull distributed random variable. Further, for forecasting, the monthly expected wind power density is extrapolated using a Gaussian process regression model.
\end{abstract}

@@

The wind speed data used in this paper was collected by the \href{https://www.dwd.de}{DWD} on \href{https://www.openstreetmap.org/?mlat=54.1750&mlon=7.8920#map=15/54.1750/7.8920}{Helgoland}, spanning the years from 1996 to 2022. The wind speed is recorded at a temporal resolution of ten-minutes\footnote{Averaged over the last ten-minute interval.}, with the anemometer being $4.38 \mathrm{m}$ above ground. Averaged over the whole measurement period, the data has a yearly mean completeness of about $91\%$. Notably, the year 2019 is missing $88\%$ of its measurements. We consider this, and a noticeable drop in the yearly wind speed mean and variance of the following years, as evidence that the measurement process was changed, and therefore discard them. Notably, the years 1996 to 1998 have a completeness of about $3\%, 85\%$ and $84\%$, respectively. For preventing a distorting effect on the Weibull distribution estimations, we also did not consider these years for our analysis, thereby improving the yearly mean completeness to $99\%$. For handling outliers, we dropped measurements exhibiting a z-score $> 8$.

@@

\subsection{Fitting of the Weibull Distribution and Trend Analysis}\label{sec:weibull}
Based on the observed wind data, we model the wind speed probability distribution both on a monthly and yearly basis.
Whilst there are several other models, like the three-parameter Weibull distribution or the normal distribution for fitting wind speed frequencies, the most prominent one is the two-parameter Weibull distribution
\cite{review, statanalysis, mohammadi2016assessing}.
For the scale parameter $\lambda >0$ and shape parameter $\beta >0$, the probability density function (PDF) and cumulative density function (CDF) are given by
\begin{align}
    \label{eq:pdf}
    p(v) &=
    \begin{cases}
        \frac{\beta}{\lambda} \left( \frac{v}{\beta}\right)^{\beta-1} 
        \exp{ \left( -\left(\frac{v}{\lambda}\right)^\beta \right)} & \text{  for } v>0 \\
        0 & \text{ else}
    \end{cases} \\
    \label{eq:cdf}
    P(v)&=
    \begin{cases}
        1- \exp{\left(-\left(\frac{v}{\lambda}\right)^\beta \right)} & \text{  for } v>0 \\
        0 & \text{ else}
    \end{cases}.
\end{align}
Via the gamma function, $\Gamma(z)= \int_{0}^{\infty}t^{z-1} \mathrm{e}^{-t}dt$, 
the $k$-th raw moment and the variance of a random variable $V$ with Weibull distribution can be expressed as follows: 
\begin{align}  
    \mathrm{E}[V^k] &= \lambda^k \Gamma \left( 1 + \frac{k}{\beta} \right) \label{eq:n-raw-moment} \\
    \mathrm{Var}[V] &= \lambda^2 \left[  \Gamma \left( 1 + \frac{2}{\beta}\right) \label{eq:var}
    - \Gamma^2 \left( 1 + \frac{1}{\beta}\right)\right].
\end{align}

@@

\subsection{Gaussian Process Regression}\label{sec:gp}
For forecasting, we employ a Gaussian process regression (GPR) model to perform a time-series regression on the monthly estimated power density and extrapolate future values. Thereby, we assume that the wind speed data $Y \in \mathbb{R}^N$ is from a latent function $f: \mathbb{R}^N \rightarrow \mathbb{R}^N$ with some added measurement noise: $y = f(x) + \epsilon$. The GPR framework consists of a Gaussian process as prior, multivariate Gaussian likelihood and Gaussian process posterior conditioned on the data 
\cite{rasmussen-williams-gp}:
\begin{align}
    \label{eq:gp_prior}
    p(f) &= \mathcal{GP}(f; 0, k), \\
    \label{eq:gp_likelihood}
    p(y \mid f) &= \mathcal{N}(Y; f, \sigma^2 I), \\
    \label{eq:gp_posterior}
    p(f \mid Y) &= \mathcal{GP}(f; m', k') \propto p(Y \mid f) \cdot p(f).
\end{align}
The resulting posterior (and prior) distribution collapses into a multivariate normal distribution when evaluated at some new locations $X' \in \mathbb{R}^M$.

@@

The least-squares linear regression of the yearly parameters can be seen in Figure \ref{fig:lin_reg}.
It shows a decrease of the parameter $\lambda$ of 
$-0.059\frac{m}{s} \frac{1}{\mathrm{yr}}$ with an RMSE of $0.42\frac{m}{s}$. 
For $\beta$, the slope is $-0.002 \frac{1}{\mathrm{yr}}$, with an RMSE of $0.07  \frac{m}{s}$.
At the $5\%$ level of the two-sided permutation test, the null hypothesis that there is no linear trend in 
the development of the parameter $\lambda$ can be rejected (p-value$=0.001$), so the observation is statistically significant. 
For the parameter $\beta$, however, the p-value is $0.383$, so we cannot reject the null hypothesis.
Assuming the trend observed in $\lambda$ and the empiric mean $\bar \beta=2.29$ for $\beta$, we derive the following trends in other quantities:
As the mean  and standard deviation are proportional to $ \lambda$, we can infer a linear trend of $-0.052 \frac{m}{s}  \frac{1}{\mathrm{yr}}$
and  $-0.024 \frac{m}{s} \frac{1}{\mathrm{yr}}$, respectively. 
%Further, the probabilities of wind speed below $ 3 \frac{m}{s}$ and above $22 \frac{m}{s}$ are especially relevant (Section \ref{sec:power-density}):
%Over the entire $22$ year period, the estimated probabilities change (in a non-linear fashion) from $6.46 \%$ to $8.15 \%$ and from $0.17 \%$ to $0.03 \%$, respectively. 
Further, under these assumptions, the mean power density 
of the mean wind speed decreases (non-linearly) from 1999 to 2018  by $30\%$.

@@

%For assessing the viability of wind turbine (parks), one might also be interested in the weekly or daily values  and, even more importantly, in a forecast. Hence, for interpolating the daily expected power density, we employ a Gaussian Process Regression model to perform a time-series regression. In addition to interpolating, we also use the same model for extrapolation and thereby making predictions about the future daily expected power density.
The first step for conducting a GP regression is to decide on a kernel $k$ for the prior \eqref{eq:gp_prior}. By analysing the plotted raw monthly expected power density, we identify a yearly periodic trend and a slight global downward trend. The former can be captured by the Exponential-Sine-Squared kernel (ES) \cite{MacKay1998IntroductionTG}. Supported by the empirical data, we choose a periodicity of $365 \, \text{days}$ and an amplitude of $450 \, \frac{\textrm{W}}{\text{m}^2}$. For the length scale, we settle for a rather small value ($< 2 \, $ days) in order to capture local patterns in the periodicity more accurately. For allowing the amplitude to slowly decay over long periods of time, we multiply the ES kernel by a very smooth (length scale $> 50 \, \text{years}$) Squared-Exponential (SE) kernel. For reproducing the global downwards trend observed in section \ref{sec:trend}, we also choose a very smooth Squared-Exponential kernel with an output scale of $100 \, \frac{\textrm{W}}{\text{m}^2}$. For accounting for the monthly midterm trend, we add a Rational-Quadratic kernel \cite{rasmussen-williams-gp} with a scale of $10 \, \frac{\textrm{W}}{\text{m}^2}$ and a length scale of one month. Finally, the additive kernel, consisting of a White-Noise (WN) and a Matern kernel (MA) \cite{abramowitz1968handbook}, is responsible for explaining measurement noise in the data and thus also prevents overfitting. Thereby, the kernel has the form
$$k = \underbrace{\text{ES} \cdot \text{SE}}_{\text{yearly trend}} + \underbrace{\text{RQ}}_\text{monthly trend} + \underbrace{\text{ES}}_\text{longterm trend} + \underbrace{\text{WN} + \text{MA}}_{\text{noise}}.$$
By drawing samples\footnote{with the mean of the data added.} $f_i \sim \mathcal{GP}(f; 0, k)$ from the prior distribution \eqref{eq:gp_prior}, we gauge how accurately our kernel choice fits to the given data. The results are shown in Figure \ref{fig:gp-samples}. We can observe that the amplitude and periodicity of some of the samples mirrors the data's in an accurate manner, supporting our choice of hyperparameters. After conditioning the Gaussian process prior on the normalised data, we obtain the posterior distribution $\mathcal{GP}(f; m', k')$ \eqref{eq:gp_posterior}. The evaluated posterior mean and standard deviation can be found in Figure \ref{fig:gp-pred}. We can observe that the mean function is following the periodicity of the ground truth without overfitting extreme power densities. Furthermore, the uncertainty of the model (standard deviation) increases gradually the further the prediction is from the data. Notably, the model also captures the global downward trend. When comparing the prediction to the test data in the years 2017 and 2018, we can observe that the prediction fits the ground truth accurately and all data points are well within one standard deviation of the mean.

@@

However, there are some shortcomings. One of the inherent limitations of the GPR is that it yields functions mapping onto $\R$, while the power density is only limited to $\R_+$. 
%Further, we conducted an interpolation of daily values based on monthly data, which does not take the inter-day variability into account. For the purpose of extrapolating daily power densities, 
%it might be advisable in further work to rely on daily data to begin with, complicating the choice of the kernel.  
Another important limitation is that our data was recorded at a height of $4.38 \mathrm{m}$, whereas the real height of wind turbines can exceed $100 \mathrm{m}$. One commonly applied rule of thumb to mitigate this issue
is multiplying the wind speeds by the factor 
$\left(\frac{h}{4.38 \mathrm{m}} \right)^{0.14}$ or the power density by its cube, 
where $h$ is the  proposed hub height of a turbine \citep{statanalysis}. However, this is far from exact, as the winds at different heights could be subject to different circulations. 
Furthermore, we did not take the direction of the wind into account and made the simplification that wind turbines are always perfectly aligned. 
Also, for equation \eqref{eq:power}, we assumed a constant air pressure rather than considering the monthly changing pressure.
Lastly, with the use of the maximum-likelihood method, we assumed the independence of measurements. While this is generally not given, we were still able to produce adequate estimations.

\section*{Contribution Statement}
David Voigt conducted the Gaussian process regression. 
Gwendolyn Neitzel was mainly concerned with the Weibull model fitting and the trend analysis.
Mohammad Fadel Berakdar focused on the data preprocessing. Alireza Yahyanejad worked on data loading.
All authors were involved in data exploration and jointly wrote the report. 
%\textcolor{magenta}{Explain here, in one sentence per person, what each group member contributed. For example, you could write: Max Mustermann collected and prepared data. Gabi Musterfrau and John Doe performed the data analysis. Jane Doe produced visualizations. All authors will jointly write the text of the report. Note that you, as a group, are collectively responsible for the report. Your contributions should be roughly equal in amount and difficulty.}
